---
title: "EasyPlotMarkdown"
author: "Alexandra Davis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#How to Use This R Markdown:

1. **Load your data**: Replace *'"data.csv"'* with the path to your dataset file. Make sure that your data frame has a column called 'value' (or replace data$value with the approriate column name).
  + Additionally, Replace the data import step              *('read.csv("data.csv")')* with the actual method you're using to load your data (such as 'read.xlsx()' for Excel files).
2. **Outlier Detection**: The IQR method is used to identify the outliers. Outliers are values that are outside the bounds of 'Q1 - 1.5 * IQR' and 'Q3 + 1.5 * IQR'.
3. **Plotting**: Ggplot is used to plot and visualize the data, showcasing the mean, Q1, median, Q3 and to highlight the outliers. 
  + If necessary, adjust the 'geom_line()' and 'geom_point()' for style or data-specific adjustments.

This R Markdown file will give you a detailed analysis of your data, including outlier detection, descriptive statistics, and a line plot to visualize these findings.

#The Code:
```{r}
#Load the needed libraries
library(ggplot2)   # For plotting
library(dplyr)     # For data manipulation
library(tidyr)     # For tidying data

#Replace "data.csv" with your path to the data file
data <- read.csv("data.csv")

#Show the first few rows of the data 
head(data)

#Summary of the dataset
summary(data)

#Calculate the quartiles and IQR
Q1 <- quantile(data$value, 0.25)
Q3 <- quantile(data$value, 0.75)
IQR_value <- Q3 - Q1

#Define the outlier boundaries
lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

#Identify the outliers
outliers <- data %>% filter(value < lower_bound | value > upper_bound)

#Display the outliers
outliers

#The descriptive statistics
mean_value <- mean(data$value, na.rm = TRUE)
median_value <- median(data$value, na.rm = TRUE)
Q1_value <- quantile(data$value, 0.25, na.rm = TRUE)
Q3_value <- quantile(data$value, 0.75, na.rm = TRUE)

mean_value
median_value
Q1_value
Q3_value

#Create a line plot with ggplot
ggplot(data, aes(x = 1:nrow(data), y = value)) +
  geom_line(color = "blue", size = 1) +
  geom_point(aes(color = (value < lower_bound | value > upper_bound)), size = 2) +  #Highlights outliers
  geom_hline(aes(yintercept = mean_value), color = "red", linetype = "dashed") +
  geom_hline(aes(yintercept = median_value), color = "green", linetype = "dashed") +
  geom_hline(aes(yintercept = Q1_value), color = "orange", linetype = "dotted") +
  geom_hline(aes(yintercept = Q3_value), color = "orange", linetype = "dotted") +
  labs(title = "Line Plot of Values with Outliers Highlighted",
       x = "Index",
       y = "Value") +
  theme_minimal() +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))  #Red points for outliers

```

